```{r, message=FALSE, echo=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
library(knitr)
library(ggthemes)

theme_set(theme_fivethirtyeight())

df_2013 <- read_csv('data/Loans_20130101to20140101_20180415T060158.csv') %>% mutate(year = 2013)
df_2014 <- read_csv('data/Loans_20140101to20150101_20180415T060226.csv') %>% mutate(year = 2014)
df_2015 <- read_csv('data/Loans_20150101to20160101_20180415T060359.csv') %>% mutate(year = 2015)
df_2017 <- read_csv('data/Loans_20170101to20180101_20180415T060911.csv') %>% mutate(year = 2017)
df_2018 <- read_csv('data/Loans_20180101toCurrent_20180415T061114.csv') %>% mutate(year = 2018)


df_2018 <- df_2018 %>% mutate(loan_default_reason = as.integer(loan_default_reason))


df_all_years <- bind_rows(df_2013, df_2014, df_2015, df_2017, df_2018)
df_all_years_raw <- df_all_years


df_all_years <- df_all_years %>% 
  mutate(lost_money = factor(ifelse(loan_status_description == 'COMPLETED' & next_payment_due_amount == 0, 0, 1)
                             , levels = c(0, 1))
         ,late_fees_flag = factor(ifelse(late_fees_paid == 0, 0, 1))
         ,has_next_payment_flag = factor(ifelse(next_payment_due_amount == 0, 0, 1), levels = c(0, 1))
         ,prosper_rating = factor(prosper_rating, levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR'))
         ,term = factor(term)
         ,year = factor(year)
         ,loan_status_description = factor(loan_status_description)
         ,service_fees_paid = service_fees_paid * -1
         ,total_paid = principal_paid + service_fees_paid + interest_paid + prosper_fees_paid + late_fees_paid
         ,total_fees_paid = service_fees_paid + interest_paid + prosper_fees_paid + late_fees_paid
         ,month_of_origination = month(origination_date))
```


# EDA

### The Question of this project: Can we predict Microlending Loan Defaults (on a website called Prosper)?

### Adrian Naranjo

<br>

### Note:
#### 1 means the loan LOST money.
#### 0 means the loan DIDN'T lose money.

<br>

#### The below shows a table of aggregate values by borrower rating, which is itself broken down into loss or not.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
aggregates <- df_all_years %>%
  group_by(lost_money, prosper_rating) %>%
  summarise(count = n()
            ,median_amt_borrowed = median(amount_borrowed)
            ,total_paid = round(median(total_paid))
            ,total_fees_paid = round(median(total_fees_paid))
            ,median_principal_paid = round(median(principal_paid))
            ,median_service_fees = round(median(service_fees_paid)))

aggregates %>% arrange(prosper_rating, lost_money) %>% kable(align=c(rep('c', ncol(.))))
```

<br>

### Counts by Year

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Counts by YEAR
df_all_years %>% 
  ggplot() +
  geom_bar(aes(year, fill = lost_money))

df_all_years %>% 
  group_by(year, lost_money) %>% 
  summarise(count = n()) %>%
  spread(key = year, value = count) %>% 
  kable(align=c(rep('c', ncol(.))))
```

<br>

### Counts by Month

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Counts by MONTH
df_all_years %>% 
  ggplot() +
  geom_bar(aes(month_of_origination, fill = lost_money))

df_all_years %>% 
  group_by(month_of_origination, lost_money) %>% 
  summarise(count = n()) %>% 
  spread(key = month_of_origination, value = count) %>% 
  kable(align=c(rep('c', ncol(.))))
```

<br>

### Counts by Term (length of the loan in months)

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Counts by TERM
df_all_years %>% 
  ggplot() +
  geom_bar(aes(term, fill = lost_money))

df_all_years %>% 
  group_by(term, lost_money) %>% 
  summarise(count = n()) %>% 
  spread(key = term, value = count) %>% 
  kable(align=c(rep('c', ncol(.))))
```

<br>

### Amount Borrowed (in USD) by Rating

#### Clearly the more you lend, the more likely you'll lose money...
```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Plots AMOUNT BORROWED by rating
ggplot(df_all_years) +
  geom_histogram(aes(amount_borrowed, binwidth = 20000), fill = "black") +
  facet_grid(lost_money ~ prosper_rating) +
  scale_x_continuous(labels = c())
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplot(df_all_years, aes(prosper_rating, amount_borrowed)) +
  geom_boxplot() +
  facet_wrap(~ lost_money)
```

<br>

### Borrower Rate by Rating

#### Unfortunately, the higher the rate, the higher the chances you'll lose money 
```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Plots BORROWER RATE by rating
ggplot(df_all_years) +
  geom_histogram(aes(borrower_rate), fill = "black") +
  facet_grid(lost_money ~ prosper_rating)
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplot(df_all_years, aes(prosper_rating, borrower_rate)) +
  geom_boxplot() +
  facet_wrap(~ lost_money)
```

<br>

### Total Fees Paid by Rating

#### Loans which lose money also have higher fees (perhaps driven by higher loan amounts?)...
```{r, message=FALSE, echo=FALSE, warning=FALSE}
# Plots TOTAL FEES PAID by rating
ggplot(df_all_years) +
  geom_histogram(aes(total_fees_paid), fill = "black") +
  facet_grid(lost_money ~ prosper_rating) +
  scale_x_continuous(labels = c())
```

```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplot(df_all_years, aes(prosper_rating, total_fees_paid)) +
  geom_boxplot() +
  facet_wrap(~ lost_money)
```